Visualizing Contingency Tables

### 2006-05-02 trunk/vcd/paper/hsc.Rnw
final Z touches on HSC chapter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@526 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

This chapter will give an overview of the state of the art 
of mosaic and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. The techniques are illustrated using three- and four-way tables.
Section~\ref{sec:concl} concludes the chapter.

@@

Throughout this section, our examples will be based on
the \data{hospital} data \cite{vcd:wing:1962} given in Tab.~\ref{tab:hospital}.

@@

\noindent The table relates the length of stay (in years) of 132
long-term schizophrenic patients in two London mental hospitals with
the frequency of visits (from relatives or friends). 
The length of stay (LOS) has been categorized into
2--9 years, 10--19 years, and more than 19 years. There are also three
categories for the visit frequency: regular (including
%FIXME: lookup regular in Wing-paper.
patients that were allowed to go home), less than monthly, and never.
Wing \cite{vcd:wing:1962} concludes from this data 
that the longer the length of stay in hospital, the less frequent the
visits, which can be seen from the column-standardized table 
(see Tab.~\ref{tab:colhospital}).

@@

\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
significantly different in the ``less than monthly'' and ``never''
strata. From the row-standardized table (see Tab.~\ref{tab:rowhospital}), 
it seems indeed that LOS is homogeneous with respect to these two visit frequency strata.

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Tab.~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Figure~\ref{fig:sievefig}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

@@

\subsection[A Note on Colors and Color Palettes]{A Note on Colors and
  Color Palettes\footnote{The printed version of the article is
    monochrome, but the electronic version uses colors. The colored
    versions of the plots are available from the code web page 
    (\url{http://statmath.wu-wien.ac.at/projects/vcd/}).}}
\label{sec:colors}

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter III-12.

@@

Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third piece of information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using less colorful (`uninteresting') colors for non-significant
results. These can again be derived using the same procedure
described in Sect.~\ref{sec:colors} but using a smaller amount of color, i.e.,
a smaller maximal chroma (e.g., 20 instead of 100).

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Tab.~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improved}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

For the visualization of areas perceptually uniform colors and color palettes
should always be used and HCL color space provides a convenient way to do so.
Shadings can be used to add information to the basic plots and to support the
analysis. Highlighting of tiles can support the analysis relationships with a 
single dependent variable. Residual-based shadings can be used for model diagnostics, e.g.,
by using diverging color palettes to code the sign and size of a residual.
In addition, the significance of test statistics can be visualized by using 
overall less colorful palettes. Using the maximum instead of the $\chi^2$ (or other) test
statistic(s) allows for data-driven cut-offs in the diverging palettes and
precise diagnostic identification of the cells in conflict with the
null hypothesis of independence.

@@

In Sect.~\ref{sec:twoway}, we have presented basic displays
for two-way tables, based on the visualization of information on 
the tables' cells, arranged in rectangular form. For multi-way tables,
mosaic plots can directly be used by simply adding further splits for each 
additional variable. For sieve and association plots,
we apply the basic idea of mosaic plots to the table itself,
i.e., simply nest the variables into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation of the multi-way table that can be visualized in %'
ways similar to a two-dimensional table.

@@

As an example, consider the well-known UCB admissions data 
\cite{vcd:Bickel+Hammel+O'Connell:1975}
on applicants to graduate school at the University of California in Berkeley for
the six largest departments in 1973 classified by admission and
gender. The `flattened' contingency table, putting department into the
columns and admit---nested in gender---into the rows, is given by
Tab.~\ref{tab:ucb}. The table aggregated over all
departments gives wrong evidence of gender bias, 
i.e., higher admission rates for male students.

@@

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression. %FIXME% mention this?
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.

@@

% As an example, consider the \data{PreSex} data
% \cite{vcd:thornes+collard:1979,vcd:gilbert:1981} on
% pre- and extra-marital sex and divorce: 494 divorced and 542 married
% people (willing to divorce) were asked two questions about their pre- and
% extramarital sexual experiences, resulting in Tab.~\ref{tab:presex}.

@@

As a four-way example, we use the punishment data \cite{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children (see Tab.~\ref{tab:punish}).      
The data consist of four variables: Attitude is a binary variable
indicating whether a person approves moderate punishment of children
(``moderate''), or refuses any punishment of children (``no'').
Memory indicates whether the person recalls having been
punished as a child. Education indicates the
highest level of education (elementary, secondary, high). Finally, 
age indicates the age group in years (15--24, 25--39, $40+$).

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))
@
\caption{Conditional mosaic plot for the \data{punishment} data. For
  each mosaic plot, memory is in the rows (first split) and attitude
  in the columns (second split).}
\label{fig:punishcotab}
\end{center}
\end{figure}

>>>>

This chapter will give an overview of the state of the art 
of mosaic and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Section~\ref{sec:twoway}, we start with the
simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Section~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. The techniques are illustrated using three- and four-way tables.
Section~\ref{sec:concl} concludes the chapter.

@@

Throughout this section, our examples will be based on
the \data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.

@@

\noindent The table relates the length of stay (in years) of 132
long-term schizophrenic patients in two London mental hospitals with
the frequency of visits (from relatives or friends). 
The length of stay (LOS) has been categorized into
2--9 years, 10--19 years, and more than 19 years. There are also three
categories for the visit frequency: regular (including
%FIXME: lookup regular in Wing-paper.
patients that were allowed to go home), less than monthly, and never.
Wing \cite{vcd:wing:1962} concludes from this data 
that the longer the length of stay in hospital, the less frequent the
visits, which can be seen from the column-standardized table 
(see Table~\ref{tab:colhospital}).

@@

\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
significantly different in the ``less than monthly'' and ``never''
strata. From the row-standardized table (see Table~\ref{tab:rowhospital}), 
it seems indeed that LOS is homogeneous with respect to these two visit frequency strata.

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Table~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Figure~\ref{fig:sievefig}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

@@

\subsection[A Note on Colors and Color Palettes]{A Note on Colors and
  Color Palettes\footnote{The printed version of the article is
    monochrome, but the electronic version uses colors. The colored
    versions of the plots are available from the Web page 
    \url{http://statmath.wu-wien.ac.at/projects/vcd/}.}}
\label{sec:colors}

@@

All plots introduced in Section~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter III-12.

@@

Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third piece of information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using less colorful (`uninteresting') colors for non-significant
results. These can again be derived using the same procedure
described in Section~\ref{sec:colors} but using a smaller amount of color, i.e.,
a smaller maximal chroma (e.g., 20 instead of 100).

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Table~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improved}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

For the visualization of areas perceptually uniform colors and color palettes
should always be used and HCL color space provides a convenient way to do so.
Shadings can be used to add information to the basic plots and to support the
analysis. Highlighting of tiles can support the analysis of relationships with a 
single dependent variable. Residual-based shadings can be used for model diagnostics, e.g.,
by using diverging color palettes to code the sign and size of a residual.
In addition, the significance of test statistics can be visualized by using 
overall less colorful palettes. Using the maximum instead of the $\chi^2$ (or other) test
statistic(s) allows for data-driven cut-offs in the diverging palettes and
precise diagnostic identification of the cells in conflict with the
null hypothesis of independence.

@@

In Section~\ref{sec:twoway}, we have presented basic displays
for two-way tables, based on the visualization of information on 
the tables' cells, arranged in rectangular form. For multi-way tables,
mosaic plots can directly be used by simply adding further splits for each 
additional variable. For sieve and association plots,
we apply the basic idea of mosaic plots to the table itself,
i.e., simply nest the variables into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation of the multi-way table that can be visualized in %'
ways similar to a two-dimensional table.

@@

As an example, consider the well-known UCB admissions data 
\cite{vcd:Bickel+Hammel+O'Connell:1975}
on applicants to graduate school at the University of California in Berkeley for
the six largest departments in 1973 classified by admission and
gender. The `flattened' contingency table, putting department into the
columns and admit---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all
departments gives wrong evidence of gender bias, 
i.e., higher admission rates for male students.

@@

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression.
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.

@@

% As an example, consider the \data{PreSex} data
% \cite{vcd:thornes+collard:1979,vcd:gilbert:1981} on
% pre- and extra-marital sex and divorce: 494 divorced and 542 married
% people (willing to divorce) were asked two questions about their pre- and
% extramarital sexual experiences, resulting in Table~\ref{tab:presex}.

@@

As a four-way example, we use the punishment data \cite{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children (see Table~\ref{tab:punish}).      
The data consist of four variables: Attitude is a binary variable
indicating whether a person approves moderate punishment of children
(``moderate''), or refuses any punishment of children (``no'').
Memory indicates whether the person recalls having been
punished as a child. Education indicates the
highest level of education (elementary, secondary, high). Finally, 
age indicates the age group in years (15--24, 25--39, $40+$).

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
set.seed(123)
cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))
@
\caption{Conditional mosaic plot for the \data{punishment} data. For
  each mosaic plot, memory is in the rows (first split) and attitude
  in the columns (second split).}
\label{fig:punishcotab}
\end{center}
\end{figure}
